\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-10-10 paper/main.tex
addressed PCF comments in abstract, motivation/significance, and impact sections

Modern research projects that rely on software are becoming
increasingly complex.  Even when a particular function or script
appears simple, the code may depend on external libraries (which in
turn may depend on other libraries), particular CPU or GPU instruction sets, and
so on.  Different versions of those external libraries (or processor
instruction sets) may ....

% IN-PROGRESS TEXT------

The same code, executed on different systems or in different computing
environments, may behave differently.  For example, when research code relies
on external libraries, the code's functionality

Reproducability and predictability are fundamental goals of nearly all
scientific undertakings.  For projects that rely on software
components, achieving these goals

% --------

Code sharing is a core component of the open science movement that has
inspired the development of a large ecosystem of tools and packages.
However, sharing code, in and of itself, does not guarantee that
others will be able to \textit{reproduce} the desired results.  For
example, research code often requires installing other software
packages that extend the implementation language's basic
functionality.  Within the Python community~\citep{vanR95}, external
packages that are published in the most popular
repositories~\citep{pypi, Anac12} are associated with version numbers
and tags that enable users to guarantee that they are installing
exactly the same code across different computing environments.
Despite that it is \textit{possible} to manually install the intended
version numbers of every dependency of a Python script or package,
doing so may cause conflicts within the user's computing environment
that interfere with the functionality of \textit{other} code.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text \textbf{Python
    scripts} (.py files) provide the
  most basic ``system'' for sharing raw code.  Scripts may reference
  external libraries, but those libraries are not automatically
  installed on other users' systems, nor is any version checking
  performed by default.  \textbf{Jupyter notebooks} (.ipynb files) comprise embedded text,
  executable code, and media (including rendered figures, code output,
  etc.).  When the \textbf{\texttt{davos} library} is imported into a Jupyter
  notebook, the notebook's functionality is extended to automatically
  install the required external libraries (at their correct versions,
  when specified).  \textbf{Virtual environments} install an isolated copy of
  Python and all required dependencies.  This typically requires
  defining a \texttt{requirements.txt} file that lists all
  dependencies (including version numbers) along with an environment
  (.yml) file that specifies how the virtual environment should be
  configured.  \textbf{Containers} provide a means of defining an isolated
  environment that includes a complete operating system (independent
  of the user's operating system), in addition to a virtual
  environment or other configurations needed to provide the necessary
  computing environment.  Containers are typically defined using
  specification files (e.g., a plain-text \texttt{dockerfile}) that
  instruct the virtualization engine regarding how to build the
  virtual environment.  \textbf{Virtual machines} provide a complete
  hardware-level simulation of the computing environment.  In addition
  to simulating specific hardware, virtual machines (typically
  specified using binary images files) must also define operating
  system-level properties of the computing environment.  Systems to
  the left of the blue vertical line entail sharing individual
  files, with no additional installation or configuration needed to
  run the target software.  Systems to the right of the red vertical
  line provide high reproducibility by supporting precise control over
  dependencies and versioning.  Notebooks enhanced using the
  \texttt{davos} library are easily shareable and require minimal
  setup costs, while also facilitating high reproducibility by
  enabling precise control over project dependencies.}
\label{fig:code-sharing}
\end{figure}

To facilitate code sharing, the Python community has developed a broad
set of approaches and tools (Fig.~\ref{fig:code-sharing}).  At one
extreme, simply publishing a set of Python scripts (.py files) may
enable others to use or gain insights into the relevant work.  Because
Python is installed by default on most modern operating systems, for
some projects this may be sufficient.  Another popular approach
entails creating JSON files, called Jupyter
notebooks~\citep{KluyEtal16}, that comprise a mix of text, executable
code, and embedded media.  Notebooks may call or import external
scripts or libraries in order to provide a more compact and readable
experience for users.  Each of these systems (Python scripts and
notebooks) provides a convenient means of sharing code, with the
caveat that they do not specify the computing environment in which the
code is executed.  Therefore the functionality of code shared using
these systems cannot be guaranteed across different computing
environments.

@@

@@

A second advanced use case is in providing a proof-of-concept of how
one can add new ``keywords'' to the Python language by leveraging the
error-handling mechanisms.  This could lead to exciting new tools
that, like \texttt{davos}, extend the Python language in useful ways.
We note that our approach to adding the \texttt{smuggle} keyword to
Python when \texttt{davos} is imported into a notebook-based
environment also has the potential to be exploited for more nefarious
purposes.  For example, a malicious user could use a similar approach
(e.g., in a different library) to substantially change a notebook's
functionality by adding new \textit{unexpected} keyword-like objects
(e.g., based around common typos).  This could lead to
difficult-to-predict changes in a notebook's behavior once the
malicious library was imported.  This highlights an important reason
why security-conscious users would be well-served to only make use of
libraries from trusted sources, or whose code is publicly available
for review.

@@

\section*{Author Contributions}
Conceptualization: PCF and JRM.  Methodology: PCF and JRM.
Implementation: PCF.  Validation: PCF.  Testing: PCF and JRM.
Writing: PCF and JRM.

>>>>

The same computer code may not behave identically under different
circumstances.  For example, when code depends on external libraries,
different versions of those libraries may function differently.  Or
when CPU or GPU instruction sets differ across machines, the same
high-level code may be compiled into different machine instructions.
Because executing identical code does not guarantee identical
outcomes, code sharing in and of itself is often insufficient for
enabling researchers to reproduce each others' work.

Within the Python~\citep{vanR95} community, external
packages that are published in the most popular
repositories~\citep{pypi, Anac12} are associated with version numbers
and tags that enable users to guarantee that they are installing
exactly the same code across different computing environments.
Despite that it is \textit{possible} to manually install the intended
version numbers of every dependency of a Python script or package,
doing so may cause conflicts within the user's computing environment
that interfere with the functionality of \textit{other} code.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{figs/shareable_code}
\caption{\small \textbf{Systems for sharing code within the Python
    ecosystem.}  From left to right: plain-text \textbf{Python
    scripts} (\texttt{.py} files) provide the most basic ``system''
  for sharing raw code.  Scripts may reference external libraries, but
  those libraries must be manually installed on other users' systems.
  Further, any checking needed to verify that the correct versions of
  those libraries were installed must also be performed manually.
  \textbf{Jupyter notebooks} (\texttt{.ipynb} files) comprise embedded
  text, executable code, and media (including rendered figures, code
  output, etc.).  When the \textbf{\texttt{davos} library} is imported
  into a Jupyter notebook, the notebook's functionality is extended to
  automatically install the required external libraries (at their
  correct versions, when specified).  \textbf{Virtual environments}
  install an isolated copy of Python and all required dependencies.
  This typically requires defining a \texttt{requirements.txt} file or
  an environment (\texttt{.yml}) file that that specifies all project
  dependencies (including version numbers of external libraries).
  \textbf{Containers} provide a means of defining an isolated
  environment that includes a complete operating system (independent
  of the user's operating system), in addition to (optionally)
  specifying a virtual environment or other configurations needed to
  provide the necessary computing environment.  Containers are
  typically defined using specification files (e.g., a plain-text
  \texttt{Dockerfile}) that instruct the virtualization engine
  regarding how to build the virtual environment.  \textbf{Virtual
    machines} provide a complete hardware-level simulation of the
  computing environment.  In addition to simulating specific hardware,
  virtual machines (typically specified using binary images files)
  must also define operating system-level properties of the computing
  environment.  Systems to the left of the blue vertical line entail
  sharing individual files, with no additional installation or
  configuration needed to run the target code.  Systems to the right
  of the red vertical line support precise control over dependencies
  and versioning.  Notebooks enhanced using the \texttt{davos} library
  are easily shareable and require minimal setup costs, while also
  facilitating high reproducibility by enabling precise control over
  project dependencies.}
\label{fig:code-sharing}
\end{figure}

Researchers, programmers, and others have developed a broad
set of approaches and tools to facilitate code sharing and
reproducible outcomes (Fig.~\ref{fig:code-sharing}).  At one
extreme, simply publishing a set of Python scripts (.py files) may
enable others to use or gain insights into the relevant work.  Because
Python is installed by default on most modern operating systems, for
some projects this may be sufficient.  Another popular approach
entails creating JSON files, called Jupyter
notebooks~\citep{KluyEtal16}, that comprise a mix of text, executable
code, and embedded media.  Notebooks may call or import external
scripts or libraries in order to provide a more compact and readable
experience for users.  Each of these systems (Python scripts and
notebooks) provides a convenient means of sharing code, with the
caveat that they do not specify the computing environment in which the
code is executed.  Therefore the functionality of code shared using
these systems cannot be guaranteed across different computing
environments.

@@

Since its initial release, \texttt{davos} has found use in a variety
of applications.  In addition to managing computing environments
for multiple ongoing research studies, \texttt{davos} is being used by
both students and instructors in programming courses such as
\href{https://github.com/ContextLab/storytelling-with-data}{\textit{Storytelling
    with Data}} \cite{Mann21b} (an open course on data science,
visualization, and communication) to simplify distributing lessons and
submitting assignments, as well as in online demos such as
{\texttt{abstract2paper}} \cite{Mann21a} (an example application of
\href{https://github.com/EleutherAI/gpt-neo}{GPT-Neo}) to share
ready-to-run code that installs dependencies automatically.

@@

A second advanced use case is in providing a proof-of-concept of how
one can add new ``keyword-like'' operators to the Python language by
leveraging notebooks' error-handling mechanisms.  This could lead to
exciting new tools that, like \texttt{davos}, extend the Python
language in useful ways within notebook-based environments.  We note that
our approach to adding the \texttt{smuggle} keyword to Python when
\texttt{davos} is imported into a notebook-based environment also has
the potential to be exploited for more nefarious purposes.  For
example, a malicious user could use a similar approach (e.g., in a
different library) to substantially change a notebook's functionality
by adding new \textit{unexpected} keyword-like objects (e.g., based
around common typos).  This could lead to difficult-to-predict changes
in a notebook's behavior once the malicious library was imported.
This highlights an important reason why security-conscious users would
be well-served to only make use of libraries from trusted sources, or
whose code is publicly available for review.

@@

\section*{Author Contributions}
\textbf{Paxton C. Fitzpatrick}: Conceptualization, Methodology,
Software, Validation, Writing - Original Draft,
Visualization. \textbf{Jeremy R. Manning}: Conceptualization,
Resources, Validation, Writing - Review \& Editing, Supervision, Funding
acquisition.
